---
title: "biostats_tricks"
output: html_document
date: "2025-11-19"
---

## T-TEST

Get T statistic (equal and unequal variance)

```{r}
# sample size
n_1 <- 20
n_2 <- 10
# Sample Mean
x_1 <- 96
x_2 <- 103
# Standard Deviation
sd_1 <- 13.76494
sd_2 <- 17.95055 

# EQUAL VARIANCE - - - - - - - - - - - - - - - - - 
# calculate the pooled variance
sp <- (sd_1^2 * (n_1 - 1) + sd_2^2 * (n_2 - 1)) /
    (n_1 + n_2 - 2)

# Use pooled variance to calculate the t-statistic
t_stat <- (x_1 - x_2) /
    sqrt(sp * (1/n_1 + 1/n_2))

# UNEQUAL VARIANCE - - - - - - - - - - - - - - - -
# Calculate the t-statistic
t_stat <- (x_1 - x_2) /
    sqrt(sd_1/n_1 + sd_2/n_2)

# Calculate the degrees of freedom (V)
V <- (sd_1^2/n_1 + sd_2^2/n_2)^2 /
    ((sd_1^2/n_1)^2 / (n_1 - 1) + (sd_2^2/n_2)^2 / (n_2 - 1))
```

Get p-value from t-statistic (this no workie)

```{r}
1-pt(-1.63, 355)

```

## ANOVA

Get F statistic

```{r}
# Number of groups
k <- 4

# Group sample sizes
n_1 <- 269
n_2 <- 53
n_3 <- 28
n_4 <- 9


# Group means
x_1 <- 115
x_2 <- 114
x_3 <- 118
x_4 <- 126


# Group standard deviations
s_1 <- 13.4
s_2 <- 10.1
s_3 <- 11.6
s_4 <- 12.2


# Grand mean is an average weighted by size of each sample
X <- (n_1*x_1 
      + n_2*x_2 
      + n_3*x_3 
      + n_4*x_4
      ) /
  (n_1 + n_2 + n_3 + n_4) 

# Between-group variance
s_b <- (n_1*(x_1 - X)^2 
        + n_2*(x_2 - X)^2 
        + n_3*(x_3 - X)^2 
        + n_4*(x_4 - X)^2
        ) /
  (k - 1) # between-group degrees of freedom

# Within-group variance
s_w <- (s_1^2 * (n_1-1) 
        + s_2^2 * (n_2-1) 
        + s_3^2 * (n_3-1) 
        + s_4^2 * (n_4-1)
        ) /
  (n_1 + n_2 + n_3 + n_4 - k) # within-group degrees of freedom
  
f_stat <- s_b / s_w

```

Get p-value from F statistic

```{r}

# quantile... thingy... Gives you the F statistic
qf(1-.01, 3, 6)

# Proportion of F density (lower tail, which is why it's subtracted from 1)
1-pf(10.86601, 3, 1494)
```

Multiple Comparisons

```{r}

# Givens
x_1 <- 118 # ex
x_2 <- 114 # cur

n_1 <- 28 # ex
n_2 <- 53 # cur

within_group_variance <- 164.0857

t_stat <- (x_1 - x_2) /
    sqrt(within_group_variance * (1/n_1 + 1/n_2))

```

# Non-Parametric Tests

## One sample / paired

Sign test

```{r}
n <- 1
# Observed number of positive signs 
D <- 1

z_stat <- (D - n/2) /
  sqrt(n/4)
```

Wilcoxon Signed-Rank

```{r}
# Smaller sum
T_sum <- 1

mu <- n*(n+1) / 4
stand_dev <- sqrt((n*(n+1)*(2*n + 1)) / 24)

z_stat <- (T_sum - mu) /
  stand_dev
```

## 2 Independent Samples

Wilcoxon Rank-Sum

```{r}
# Sample sizes
n_s <- 1 # Smallest sum of ranks
n_l <- 2 # Largest sum of ranks

# Smaller sum
W_sum <- 1

mu <- (n_s*(n_s + n_l + 1)) / 2
stand_dev <- sqrt((n_1*n_2*(n_s + n_l + 1)) / 12)
  
z_stat <- (W_sum - mu) /
  stand_dev
```

# Chi-Square

Get test statistic

```{r}
# Quantile thingy - use this to get the critical value.
qchisq(1-.05, df=7)

# We subtract it from 1 because it gives us the left side of the distribution
1-pchisq(3.722, df=1)

```



# Probability Mass Function (PMF)

Given sample size n, k groups, and an event probability of p, what is the probability that k successes will happen?

```{r}
k <- c(0, 1, 2, 3, 4) # k can be a single number or a vector
n <- 40
p <- .15

n_choose_k <- factorial(n) /
    (factorial(k) * factorial(n - k))

probability <- n_choose_k * p^k * (1 - p)^(n - k)

# Sum them all together if you used multiple ks.
sum(probability)
```

# Wrangle a 2x2 table

```{r}
# Get a table that you can use for making Odds Ratios ready for input into stata
two_by_two <- dat %>% 
    group_by(
        case = ifelse(organism_or_disease_identified == 'Rickettsia africae', 'a_rickettsia', 'b_not_rickettsia'),
        exposure = ifelse(sex == 'Female', 'Female', 'Male')
        ) %>% 
    summarise(count = n())

two_by_two$count

# This also works
table(x, y)
```



# Chi-squared test
```{r}
# Wrangle into a rxc table (more complicated way)
# chi_table <- dat %>% 
#     # Remove the nulls
#     filter(!is.na(food_insecurity_didnt_last), !is.na(rejection)) %>% 
#     # Get counts for each group
#     group_by(food_insecurity_didnt_last, rejection) %>% 
#     summarise(count = n()) %>% 
#     # Pivot it into a rxc format 
#     pivot_wider(names_from = food_insecurity_didnt_last, 
#                 values_from = count) %>%
#     # Make sure the row variable gets stored as column names rather than values (will mess up structure and datatype if not included)
#     column_to_rownames('rejection') %>% 
#     as.matrix()

# The easiest way to do a chi-square test
library(janitor)

# Chi-square
chisq.test(tabyl(dataframe, x, y))

# Check expected counts
chi_sqr$expected

# If expected counts are less than 5, do a non-parametric version of the test with simulate p-value argument. 
# Fisher's exact
fisher.test(tabyl(dataframe, x, y), simulate.p.value = TRUE)

```


## Cronbach's Alpha
This test can be used to see if two similar survey questions measure the same thing. This is useful when checking questions for internal consistency reliability.
Note that this function seems to need the data to be numeric. Consider recoding ordinal variables to integers.
Read more here: https://statisticsbyjim.com/basics/cronbachs-alpha/
```{r}
library(psych)

# Make a dataframe for only the columns that are supposed to be measuring the same thing
trees <- Orange %>% 
    select(age, circumference)

alpha(trees)


```

## Check assumptions for all kinds of models
```{r}
library(easystats)

mylm <- lm(mpg ~ wt + hp + disp, mtcars)

check_model(mylm)

```

