---
title: "data wrangling secrets"
output: html_document
format: 
  html:
    code-fold: true
execute:
  warning: false
---

```{r}
library(tidyverse)
library(pander)
library(nycflights13)
library(tidyquant)
library(car)
```

SELECT COLUMNS BY REGEXP
```{r}
select(matches('your regex pattern here'))
```


LUBRIDATE
```{r}

d1 <- 'January 21st, 2017'
d2 <- '2012-01-01'
d3 <- 'March 2016 - 23'

# Convert string to a date
mdy(d1) # mdy is for mm/dd/yyyy
ymd(d2)
myd(d3)

make_date(1977, 4, 25)

jan31 <- ymd("2013-01-31")
jan31_t <- ymd_hms("2013-01-31 06:35:27")
# Extract the year: 
year(jan31_t)
# Rounds to the hour. 
round(jan31_t, unit = 'hour')

# What weekday? 
date <- 'March 23, 2016'
date <- mdy(date)
# wday retreives the weekday. label = TRUE shows the levels so you can see which number is which weekday. 
wday(date, label = TRUE)
# What day of the year?
yday(date)

# How to parse dates from Excel's format.
# It's supposed to be the number of days since Jan 1 1900, but it will result in dates
# That are 2 days off, so use '1899-12-30' as the origin.
RecordedDate = as.Date(as.numeric(RecordedDate), origin = '1899-12-30')

# Datediff in years
age_post = interval(earlier_date, later_date) %/% years(1) # integer division by years

  mutate(
    date = as_date(date, format='%m/%d/%Y'), # one option
    date = mdy(date) # does the same thing (will work even if the delimiter is / instead of -)
    date = as_date(str_sub(date, 0, 10)) # Will work for first 9/10 letters in qualtrics dates
)


  
```
Use the format argument to specify the order in which your data comes (not the order you want it to be, which will be the consistent lubridate format)

PAIRS PLOT
```{r}

# Make sure you use this argument when inputting data:
dat <- read.csv('path', stringsAsFactors = TRUE)
# It saves the strings as factors so they can be pairs plotted.

pairs(dat, panel = panel.smooth)
# panel argument adds red regression line for some reason

# How to do it on a series of columns [rows, columns]
pairs(pairs_dat[,1:15], panel = panel.smooth)

# Get sample for pairs plot
sample <- slice_sample(dat, n = 100)



```

SUBSETTING/SLICING IN BASE R
```{r}
df[1:3] # This gives you the first 3 columns
df[,1:3] # the first 3 columns
df[1:3,] # the first 3 rows
df[-3] # everything but column 3
df[-(1:9)] # everything but columns 1-9
```

REARRANGE COLUMN ORDER
```{r}
df %>% relocate(col) # move col to front
df %>% relocate(col, .after = last_col()) # move col to end

df %>% relocate(col1, .after = col2) # move col after col2 (also u can use "before")
```


## MANY-TO-MANY SOLUTIONS
If your variable is a column full of comma-separated lists and you want a row for each item (similar to a left-join, where you have to split the list from a column) use the separate_longer_delim column.
turns for example diabetes, stroke into one row for diabetes and one row for stroke (ID is duplicated)
```{r}
split_dat <- dat %>% 
    separate_longer_delim(
        cols = chronic_conditions,
        delim = ','
    )

# Use regexp in the delim argument if ur feelin frisky
split_dat <- dat %>% 
    separate_longer_delim(
        cols = chronic_conditions,
        delim = regex(',(?![:space:])') # Separates  by commas not followed by a space
    )
```


COUNT THE NULLS IN EACH VARIABLE IN THE DATASET
```{r}
colSums(is.na(iris))
```



GET A SMALL SAMPLE OF ROWS FROM A DATAFRAME AND USE PAIRS PLOT ON ONLY SOME COLUMNS
```{r}
dat_sample <- slice_sample(iris, n = 4)

# Square brackets index the dataset. The first slot (empty here) is the rows, the second slot is the columns (indexed by number)
pairs(sample[,15:28], panel = panel.smooth)

```


CALCULATE PERCENTAGES ACROSS GROUPS
```{r}
library(janitor) # for tabyl function

# 1 VAR
tabyl(snp$org_type)
# cols of this table are easily modified
tabyl(snp$org_type) %>% mutate(percent = round(percent, 2))
# Tabulate by two variables (nxn table)
tabyl(data, grant_year, site)


# 2 VARS
# The percentage of clients who said yes or no in their econ group
perc_df <- dat %>% 
    group_by(trust, user_language) %>% 
    count() %>% 
    # Groupby again to get percentages of each econ_group
    group_by(trust) %>% 
    mutate(perc = n/sum(n)*100)

# 3 VARS: confidence group is the group of interest.
perc_df <- plot_df %>% 
    group_by(cohort_label, pre_or_post, confidence) %>% 
    count() %>% 
    # Groupby again to get percentages of each econ_group
    group_by(cohort_label, pre_or_post) %>% 
    mutate(perc = n/sum(n)*100)


```




REMOVE A ROW BY ITS ROW NUMBER (even if there's no ID column)

```{r}

outlierless_davis <- Davis %>% 
  slice(-12)

```

The negative sign indicates a row removal. Remove the negative sign to make the dataframe only row 12. Multiple rows can be selected with a range

```{r}

outlierless_davis <- Davis %>% 
  slice(12:14)

# Negative is used here too in order to remove the group of rows.
outlierless_davis <- Davis %>% 
  slice(-(12:14))

```

For other options, investigate slice_min, slice_max, slice_head, etc


RUN A FUNCTION ON EVERY THING IN A VECTOR
```{r}
add_3 <- function(x) {
  y =  x + 3
  return(y)
}

iris %>% 
  mutate(
    # map() parameters: vector you apply function to, function name
    added_3 = map(Sepal.Width, add_3)
  )


```
map_dbl() can return the results as a numeric vector. Here's the ones for the other datatypes: 
map()	list
map_chr()	character vector
map_dbl()	double (numeric) vector
map_dfc()	data frame (output column binded)
map_dfr()	data frame (output row binded)
map_int()	integer vector
map_lgl()	logical vector
walk()	returns the input invisibly (used to trigger side effects)

no workie :(
```{r}

deaths <- ldeaths %>%
    map(t) %>%
    map_dfr(as_tibble)

deaths
```


RUN A FUNCTION ON MULTIPLE COLUMNS
```{r}
# Use the across() function (args: vector of columns, lambda functions to use
        across(.cols = c(emptiness, people_around, rejection),
               .fns = ~ recode(.x, # .x is the parameter for whatever's in the .cols vector. fns stands for functions
                               'Yes' = 0, 
                               'More or less' = 1, 
                               'No' = 2
                               )
               )
```




REMOVE ALL NULLS pipe it 
```{r}
na.omit()
```

REMOVE SOME NULLS
```{r}
drop_na(column_name)
```

If all else fails, try filtering everthing but NA.


DROP COLUMNS
```{r}
noodle <- mpg %>% 
  # You can also use c() vector notation to delete multiple columns. 
  select(!cyl)
```


UNGROUP WILL KEEP YOU FROM LOSING ALL THOSE COLUMNS HALLELEUJA    
```{r}

```


GRAB TOP 3 
```{r}

iris <- iris %>% 
  slice_max(n = 3, order_by = Sepal.Width)

```


TIDYQUANT GRAB STONKS DATA
```{r}
# Pulls data straight from online!
price_dat <- tq_get(c('CPIAUCSL'),
                    get = 'economic.data',
                    from = '2020-01-01')
```



JOINS
```{r}
# A default left join will match it up by all columns with the same name. However, with the flights dataset, there are 2 columns with the name 'year' that have different values. They don't combine correctly, so we have to specify that we only have to join by tailnum. 
planes_flights <- flights %>% 
  left_join(planes, by = c('tailnum'))

# This works when the ids have the same name too
falls <- right_join(
    pre, post, by = join_by(ID)
)

```


PIVOTING
```{r}

classes <- read_csv("https://byuistats.github.io/M335/data/rcw_reshape.csv")

# Convert it to wide
classes_wide <- classes %>% 
  pivot_wider(names_from = Department, values_from = Count)
  # names_from: where should we grab the column names from?
  # values_from: where should we grab the values from? 

# Convert it to long
classes_long <- classes_wide %>% 
  # Specify what you want to stick into a column? 
    pivot_longer(cols = Chem:PHY, names_to = 'Department', values_to = 'Count') 
  # Colon says take the columns from chem to phy.
  # names_to: what do we call the new column stiched together from the ones you split it from.

# Multi-column pivot_longer
    pivot_longer(
        # Select analysis columns only
        cols = c(matches('fear_p'), matches('capable'), matches('concern')&!matches('move'),
        names_to = c('metric', 'pre_or_post'), # New names columns
        names_sep = '_(?=pre|post)', # Split on underscores followed by "pre" or "post"
        values_to = 'response' # New value column
    )
# Note how names sep tells it where to split. It's regexp compatible.
```

PIVOTING A MANY-TO-MANY VARIABLE INTO ROWS FOR SUMMARIES OF EACH ITEM
```{r}
# Longer dat has a row for every condition. Participants are duplicated for each new condition they have.
# Like a left join where left = condition, right = participant row
longer_dat <- dat %>% 
    separate_longer_delim(
        cols = chronic_conditions,
        delim = regex(',(?![:space:])') # Delimiters are commas followed by a space
    )

# Now use group by to make relevant summaries for each condition
summary_df <- longer_dat %>% 
    filter(
        # Remove the duplicated ids
        !(participant_id %in% duplicated_ids)
        ) %>% 
    group_by(chronic_conditions) %>% 
    count() 
```



HOw to slice ??? Need to do more research
```{r}
library(Lahman)

Salaries %>%
  group_by(playerID) %>% 
  arrange(desc(salary)) %>% 
  # Gives you the first line of each player
  slice(1)


last_college <- CollegePlaying %>% 
  left_join(Schools) %>% 
  group_by(playerID) %>% 
  arrange(desc(yearID)) %>% 
  slice(1) %>% 
  filter(state == 'UT') %>% 
  select(-yearID)
```



RENAME COLUMNS
```{r}
# All at once
colnames(iris) = c('sepal_length', 'sepal_width', 'petal_length', 'petal_width', 'species')

# Only some
my_data_renamed <- my_data %>%
    rename(new_name = old_name)
```


BETWEEN - do vector values fall in a range? 
```{r eval=FALSE}
iriss <- iris %>% filter(iris, between(Sepal.Width, 3.5, 4))
```


APPROXIMATE IN CONDITIONAL STATEMENTS
```{r}
# Returns FALSE
sqrt(2)^2 == 2

# Returns TRUE
near(sqrt(2)^2, 2)
```


Stick na.rm = TRUE in as an argument of the mean() function.
```{r}

table <- iris %>% 
  group_by(Species) %>% 
  summarise(mean = mean(Sepal.Length, na.rm = TRUE))

pander(table)

```


arrange() orders your table - this puts lower sepal length columns first.
```{r}
practice <- iris %>% 
  arrange(Sepal.Length) %>% 
  group_by(Species) %>% 
  mutate(
    ratio = Sepal.Width / Sepal.Length,
    mean = mean(ratio),
    'standard deviation' = sd(ratio)) %>% 
  select(Species, ratio)

head(practice, n=7)

```


ARRANGE but in descending order
```{r}
stuff <- iris %>% 
  arrange(desc(Sepal.Width))
pander(stuff)
```



### Tools for importing hard to read data:  

- Use a temp file (read data from a site without keeping it)
```{r}
library(readxl)
library(downloader)

# Make a temporary directory and download data to that directory.
jelly <- tempfile()
download(
  'https://byuistats.github.io/M335/data/cpi_food_uk/cpiaverageprices201910.xlsx',
   destfile = jelly,
   mode = "wb")

# Read in data.
bread <- read_excel(jelly)

# Make new column names. 
colnames(bread) <- c('key', 'purchase', 'price')

# Drop first column with the akward date.
bread <- bread %>% 
  drop_na()



```


- SWISS ARMY KNIFE FOR EVERYTHING (rio library)  
It can recognize file extentions.
```{r}
library(rio)

# Add tibble isn't necessary, it converts the data into a tibble. Also you can directly rename the column names! Yeehaw
dat <- import(
  "https://byuistats.github.io/M335/data/cpi_food_uk/cpiaverageprices201911.xlsx", 
  skip = 2, 
  col_names = c('key', 'purchase', 'price')) %>% as_tibble()

view(dat)
```


- read_lines() function is rough but it can handle pretty much anything. Reads in everything as a string. 

HOW TO UNZIP this converts it to an sf file, try to find the one where it just goes straight to a dataframe. 
```{r}


pacman::p_load(tidyverse, sf, downloader, fs)

river_path <- 'https://byuistats.github.io/M335/data/water.zip'

# Make 2 temp files
df <- tempfile()
uf <- tempfile()

# Download from river path
download(river_path, df, mode = 'wb')
unzip(df, exdir = uf) # exdir exports to a directory uf. 

# After the data's unpacked, read it in as an sf object.
output <- read_sf(uf)

# Delete the temp files because they can be heccin chonkers and u wanna save space
file_delete(df); dir_delete(uf)

head(output)


```





# {.tabset}

DATATYPES INFO
```{r}
# Atomic vectors
numeric_vector <- c(1, 2, 3, 10:15)
char_vector <- c('raven', 'moose', 'lammergeier')

# Combining the two turns everything into strings. Even if you pipe it into as.numeric(), the strings will be turned null
c(numeric_vector, char_vector)

# Lists
# If you make both numbers and characters part of a list, it'll give you a list of lists. 
list(numeric_vector, char_vector)

# You can fill it with junk!
my_list <- list(c(1,2,3,4,5), "cat", "dog", c("mouse", "rat"), list(c("elephant", "lion", "giraffe", "zebra"), c(100:105)))

# Single brackets for single element of atomic vectors, double brackets for list
my_list[[1]][2]
# grab u a rat
my_list[[4]][2]
# grab u a zebra
my_list[[5]][[1]][4]

# Here's the better way. It goes by sequence, regardless of lists or vectors. tidyverse 
my_list %>% pluck(5, 1, 4)


# Grab the films Luke Skywalker is in, 2 diff methods
starwars %>% pluck('films', 1)
starwars[['films']][1]

```

MAKE A SIMPLE MATRIX FOR STUFF LIKE CHI^2
```{r}
chi_matrix <- matrix(c(9, 44, 13, 10, 11, 52, 23, 22, 9, 41, 12, 27), ncol = 4, byrow = TRUE)

```

