---
title: "ggplot secrets"
output: html_document
---

```{r}
library(tidyverse)

```

REMOVE NULLS FROM SHOWING UP IN PLOT
```{r}
ggplot(volun, aes(x = friendly, fill = friendly), na.rm = FALSE) +

```
Use the na.rm argument!!!


REMOVE SCIENTIFIC NOTATION
```{r}
library(scales)

options(scipen = 999)
```

SIDE BY SIDE BAR PLOTS
```{r}
ggplot(dat, aes(x = sex, fill = rickettsia_or_no)) +
    geom_bar(position = 'dodge')
```

PERCENTAGES BAR PLOT
```{r}
ggplot(RQ1, aes(x = factor(round_final), fill = attendance_frequency)) +
  geom_bar(position = "fill") # Position = 'fill' is what does that percentage plot automatically!
```

GGPLOT AUTO THEMES
```{r}
theme_classic()
theme_minimal()
theme_bw()
```



MODIFY AXIS TICK INTERVALS
```{r}
# Use the breaks label in the scale continuous function
ggplot(iris, aes(x = Species, y = Sepal.Width / Sepal.Length)) +
    geom_point() +
    scale_y_continuous(breaks = seq(.3, .8, by = .01))

# When you're working with dates:
+ scale_x_date(breaks = function(x) seq.Date(from = min(x), 
                                               to = max(x), 
                                               by = "1 month")) +

# For date
scale_x_date(date_breaks = '3 months', date_labels = '%Y%b') +

```


FORMAT AXIS AS PERCENTAGES
```{r}
ggplot(iris, aes(x = Species, y = Sepal.Width / Sepal.Length)) +
  geom_point() +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) 
  
```


REFORMAT AXIS LABELS INTO DOLLAR FORMAT 
```{r}
ggplot(mtcars, aes(x = disp)) +
  geom_bar() +
  scale_x_continuous(labels = dollar_format()) 

```


ADD UNITS TO AXIS LABELS
```{r}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) +
  geom_point() +
  # Give unit of kg to each one!
  scale_y_continuous(labels = unit_format(unit = 'ft')) +
  scale_x_continuous(labels = unit_format(unit = 'ft')) 

```


MAKE GEOM_LINES OF MULTIPLE LINES (time series, etc)
```{r}

ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, group_by = Species)) +
  geom_line()

```



GEOM COL FOR X/Y BARPLOTS
```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length)) +
  geom_col()
```


ADD LABELS TO ELEMENTS
```{r}

p <- ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, color= Species)) +
  geom_point()


# directlabels method
library(directlabels)
p + geom_dl(method = 'smart.grid', mapping = aes(label = Species))

# Label every hecking point!
p + geom_text(aes(label = Species))

# Adds a white box behind everything.
p + geom_label(aes(label = Species))

# Restrict labels to avoid overlapping, add lines for those that are far away.
library(ggrepel)
p + geom_text_repel(aes(label = Species))


# Add labels to a stacked percentage bar chart
# Manually calculate percentages
perc_df <- plot_df %>% 
    group_by(cohort_label, pre_or_post, confidence) %>% 
    count() %>% 
    # Groupby again to get percentages of each econ_group
    group_by(cohort_label, pre_or_post) %>% 
    mutate(perc = n/sum(n)*100)

ggplot(perc_df, aes(x = pre_or_post, y = perc, fill = confidence)) +
    geom_col() +
    geom_text(
        aes(label = paste0(round(perc, 1), '%')),
        position = position_stack(vjust = .5),
        color = 'white'
    ) +

```



DELETE AN AXIS LABEL
```{r}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point() +
  theme(
    axis.title.x = element_blank()
  )
```

CUSTOM THEME THAT'S NICE AND MINIMAL
```{r}
# Just horizontal lines, no ticks or panel
theme(
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_line(color = 'gray85'),
    panel.grid.major.x = element_blank()
)
```


2 GGPLOTS ON ONE LINE
```{r}
library(gridExtra)

plot1 <- ggplot(mtcars, aes(x=mpg)) +
  geom_histogram()

plot2 <- ggplot(mtcars, aes(x=qsec)) +
  geom_histogram()

grid.arrange(plot1, plot2, nrow=1)
```


REORDER
```{r, reorder}
# This organizes each one by the median! COOL BEANS
ggplot(mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy))

# Or you can do it by number of observations in each level
ggplot(mpg, aes(x=fct_infreq(manufacturer))) +
  geom_bar()

# By value.
ggplot(mpg, aes(x = fct_reorder(manufacturer, hwy), y = hwy)) +
  geom_col()

# This doesn't quite work, see reducing_gun_deaths for a working example
weather <- Weather %>% 
 mutate(
    season = case_when(
    month %in% c(12, 1, 2) ~ 'winter',
    month %in% c(3, 4, 5) ~ 'spring',
    month %in% c(6, 7, 8) ~ 'summer',
    month %in% c(9, 10, 11) ~ 'fall'
    ),
    season = factor(season, levels = c('winter', 'spring', 'summer', 'fall')))

ggplot(Weather, aes(x=season, y=high_temp)) +
  geom_col()
 

# Reorder legend independent of plotting order!
guides(fill = guide_legend(reverse = TRUE)) +

```


SCALE COLOR MANUAL
```{r}
ggplot(data = iris, mapping = aes(x=Sepal.Width, 
                                  y = Sepal.Length, 
                                  color = Species,
                                  shape = Species)) +
  geom_point() +
  # You can specify which color goes to which, but if you don't it'll organize the colors alphabetically by factor.
  scale_color_manual(values = c(setosa = 'purple4', versicolor = 'orange3', virginica = 'skyblue'))
# Use scale color or scale fill depending on what's being mapped.

# Use the labels parameter of the scale_manual functions to rename the elements in the legend. 
```


COLOR WITH A GRADIENT
This only works with continuous variables. Don't forget to specify the mapping for the colored variable in the ggplot aes.
```{r}
ggplot(mtcars, aes(x=mpg, y=qsec, color=wt)) +
  geom_point() +
  scale_color_gradient2(low = 'olivedrab', mid='gold', high='ivory')

```

USE VIRIDIS COLOR PALETTES
https://jmsallan.netlify.app/blog/the-viridis-palettes/
```{r}

# discrete
scale_fill_viridis_d() # discrete
scale_fill_viridis_b() # binned


```


HIDE LEGEND
```{r}
# Remove all legend
ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species)) +
  geom_boxplot() +
  theme(legend.position = "none")

# Remove a specific legend
ggplot(iris, aes(x=Species, y=Sepal.Length, alpha=.8)) +
  geom_boxplot() +
  scale_alpha(guide = 'none')

ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species, alpha=.8)) +
  geom_boxplot() +
  guides(fill = 'none')

```

REMOVE A SPECIFIC LEGEND must be applied to a continuous variable
```{r}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, color=Sepal.Length)) +
  geom_point() +
  scale_color_continuous(guide = 'none') 

```

PUT LEGEND AT BOTTOM
```{r}
    theme(
        legend.position = "bottom", # puts it at bottom
        legend.direction = 'vertical' # makes em stack vertical
        )
```


CHANGE FONT OF WHOLE PLOT
```{r}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point() +
  theme(
    text = element_text(family = 'serif')
    ) +
  labs(title = 'Iris')
```

SAVE A PLOT WITH SIZE & RESOLUTION CONSIDERATIONS
```{r}
ggsave(
    filename = 'plot.png',
    plot = plot,
    dpi = 300
)
```

